Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-29 paper/gd1.tex

% # style notes
% - it is Cram\'er--Rao not Cram\'er-Rao. And yet Fisher-matrix not Fisher--matrix.

@@

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\textsl{Pan-STARRS}}
\newcommand{\DR}{\acronym{DR2}}
\newcommand{\msun}{\textrm{M}_\odot}
\newcommand{\kpc}{\textrm{kpc}}
\newcommand{\mag}{\textrm{mag}}
\newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\masyr}{\ensuremath{\textrm{mas}~\textrm{yr}^{-1}}}
\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
\newcommand{\article}{\textsl{Letter}}

@@

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of matter in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the distribution of dark matter.
We present a sample of highly-probable members of the longest cold stream, GD-1,
selected to have small parallaxes and retrograde proper motions using data from
the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor
population at a distance of $\sim10~\kpc$.
This selection extended the stream by $10^\circ$(?), revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
Our model of GD-1 indicates that the stream crosses the Galactic plane at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
\end{abstract}

@@

\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
(\citealt{Carlberg:2013}).
More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}):
[summarize their conclusions].

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{stream-selection.pdf}
\end{center}
\caption{
    \todo{AB}
\label{fig:selection}
}
\end{figure}

@@

All of the clearly identifiable portions of the GD-1 stream are located at high
Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
dust extinction or variations in extinction along the stream.
\figurename~\ref{fig:sfd-cmd} (top panel) shows the $V$-band extinction in the
region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
extinction map (\cite{Schlegel:1998}; hereafter SFD).
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\mag$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
0.03~\textrm{mag}$.
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{sfd-cmd.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:sfd-cmd}
}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{track-and-model.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:track-and-model}
}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{spur.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:spur}
}
\end{figure}

@@

\acknowledgements{
It is a pleasure to thank
Vasily Belokurov,
Andrew R. Casey,
Marla Geha,
David W. Hogg,
Kathryn V. Johnston,
Sergey Koposov,
Mariangela Lisanti,
Edward Schlafly,
and David N. Spergel for useful discussions and feedback.

>>>>

@@

% typography
\setlength{\parindent}{1.\baselineskip}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\pans}{\textsl{Pan-STARRS}}
\newcommand{\DR}{\acronym{DR2}}
\newcommand{\msun}{\textrm{M}_\odot}
\newcommand{\kpc}{\textrm{kpc}}
% \newcommand{\mag}{\textrm{mag}}
\newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\masyr}{\ensuremath{\textrm{mas}~\textrm{yr}^{-1}}}
\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
\newcommand{\article}{\textsl{Letter}}

@@

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold streams
of stars that are extremely valuable tracers of the large- and small-scale
distribution of matter in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the distribution of dark matter.
We present a sample of highly-probable members of the longest cold stream, GD-1,
selected to have small parallaxes and retrograde proper motions using data from
the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor
population at a distance of $\sim10~\kpc$.
This selection extended the stream by $20^\circ$, revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
Using the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
Our model of GD-1 indicates that the stream crosses the Galactic plane at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
\end{abstract}

@@

\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
(\citealt{Carlberg:2013}).
More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2017}):
[summarize their conclusions].

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{gd1_sample.pdf}
\end{center}
\caption{
    \todo{AB}
\label{fig:selection}
}
\end{figure}

@@

All of the clearly identifiable portions of the GD-1 stream are located at high
Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
dust extinction or variations in extinction along the stream.
\figurename~\ref{fig:sfd-cmd} (top panel) shows the $V$-band extinction in the
region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
extinction map (\cite{Schlegel:1998}; hereafter SFD).
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
0.03~\textrm{mag}$.
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable.

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{sfd-cmd.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:sfd-cmd}
}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{track-and-model.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:track-and-model}
}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{spur.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:spur}
}
\end{figure}

@@

\acknowledgements{
It is a pleasure to thank
Vasily Belokurov,
Andrew R. Casey,
Marla Geha,
David W. Hogg,
Benjamin Johnson,
Kathryn V. Johnston,
Sergey Koposov,
Mariangela Lisanti,
Edward Schlafly,
and David N. Spergel for useful discussions and feedback.
